# Preparing the Precipitation Database {#precipdata}

The example workflow presented in this document relies on daily precipitation observations from the Global Historical Climatology Network (GHCN; Menne et al. 2012). Download the station inventory for the GHCN-daily dataset ([GHCNd](https://www.ncei.noaa.gov/pub/data/ghcn/daily/ghcnd-inventory.txt)). Pull the data into a spreadsheet and save as a csv named `ghcnd-inventory.csv`. Give the columns the following names: `ghcnd_id`, `lat`, `lon`, `varid`, `startyr`, `endyr`. The `startyr` and `endyr` columns aren’t used in this analysis. Note that you need to define two sub-directories within the MSP analysis directory named `Data` and `input_files`. `Data` will hold the GHCNd station data you are going to download; `input_files` will save the resulting annual maxima and metadata, along with all other required inputs for the MSP analysis.

```{r}
df <- read.csv("ghcnd-inventory.csv")
head(df)
```
## Download GHCNd stations {#download}

Plot the stations in GIS with your basin shapefile and pertinent raster data to assist in the definition of a study region surrounding the basin. Consult a hydrometeorologist (such as John England, Samantha Hartke, or yours truly) if you need help defining a bounding box. Once you have the bounding box defined, create a subset of the station inventory and download the data for those stations using the following R code (`0a_download_ghcnd.R`):

```{r eval=FALSE}
library(curl)

# set your working directory to the folder you want to save all the GHCNd data
setwd("C:/Path/To/MSP/Analysis/Data/")

# define the output directory - this will be the input to the MSP, call it "input_data"
idir = "C:/Path/To/MSP/Analysis/input_data/"

# read in the inventory file, do NOT put the inventory file in your data folder. 
df.m = read.csv("C:/Path/To/MSP/Analysis/ghcnd-inventory.csv")  

# take subset of inventory file; constrain to bounding box and only PRCP stations
# change your bounding box as required!
df.sub = subset(df.m, lon > -85 & lon < -82.8 & lat > 33.5 & lat < 35.5)
df.sub = subset(df.sub, varid == "PRCP")

# set up a progress bar to gauge how long it will take
pb = txtProgressBar(min=0,max=nrow(df.sub),style=3)

# loop through each row of the subset and attempts to download a file for that station.
# destfile is saving it to your current working directory, use getwd() to check where
for(i in 1:nrow(df.sub)){
  setTxtProgressBar(pb,i)
  curl_download(url=paste0("https://www.ncei.noaa.gov/data/global-historical-climatology-network-daily/access/",df.sub$ghcnd_id[i],".csv"),
                destfile=paste0(df.sub$ghcnd_id[i],".csv"))
}

# save the subset file for later use. do NOT save the subset file to the data folder
write.csv(df.sub, paste0(idir, "GHCNd-metadata.csv"), row.names=FALSE)
```

## Combine and Organize GHCNd Data

After downloading all the data for the stations in the bounding box, that data needs to be extracted and put into a (more) concise format. The first step is to loop through all the files to define the earliest and latest date to set up a matrix that will hold all sites (rows) and observations (columns) for every day from the earliest to the latest date. This will be a huge matrix! Use the following R code for this task (`0b_ghcnd_dataset.R`):

```{r eval=FALSE}
library(lubridate)
library(plyr)

# read in that subset we made from the station inventory and bounding box
df.meta = read.csv(paste0(idir,"GHCNd-metadata.csv"))

# fn is a list of all filenames; fns is the same with “.csv” removed (just site ID)
fn = list.files(pattern=".csv")
fns = unlist(strsplit(fn, split=".csv"))

# start with some arbitrary min and max dates – update as needed for your analysis
mind = ymd("2005-01-01")
maxd = ymd("2005-12-31")

# loop through and update ‘mind’ and ‘maxd’ variables as needed
pb = txtProgressBar(min=0, max=length(fn), style=3)
for(i in 1:length(fn)){
  setTxtProgressBar(pb,i)
  
  df = read.csv(fn[i])
  if(min(ymd(df$DATE)) < mind) mind = min(ymd(df$DATE))
  if(max(ymd(df$DATE)) > maxd) maxd = max(ymd(df$DATE))
}

# define vector of all dates, and the matrix that will hold all the data
full_dates = data.frame(DATE=seq.Date(from=ymd(mind), to=ymd(maxd), by="day"))
df.daily = matrix(NA, nrow=nrow(full_dates), ncol=length(fn))

# loop through and populate that matrix
pb = txtProgressBar(min=0,max=ncol(df.daily),style=3)
for(i in 1:ncol(df.daily)){
  setTxtProgressBar(pb,i)
  tmp = read.csv(fn[i])
  sub = subset(tmp, select=c("DATE","PRCP"))
  sub$DATE = ymd(sub$DATE)
  tmp.full = join(full_dates, sub, by="DATE")
  df.daily[,i] = tmp.full$PRCP
}

# put the station id in the column name
colnames(df.daily) = fns

# save it. do NOT save the data matrix to the data folder
write.csv(df.daily, paste0(idir,"ghcn-daily.csv"), row.names=FALSE)
```

## Calculate Annual Maxima {#anmax}

Now that the daily precipitation observations have been put into the big data matrix, it’s time to process those annual maxima. For this approach we are looking at the calendar year annual maximum n-day precipitation observation at each site independently. Use the following R code to create the n-day annual maxima (`0c_calc_anmax.R`):

```{r eval=FALSE}
library(zoo)
library(plyr)

# define output matrix (nrow = years, ncol = stations)
anmax.out = matrix(NA, nrow=length(unique(year(ymd(full_dates$DATE)))), ncol=length(fn))

# 2-day critical duration (currently only works for daily scale, we're using GHCNd after all!)
dur = 2  

# loop through each row of df.daily and calculate annual maxima
pb = txtProgressBar(min=0, max=ncol(df.daily), style=3)
for(i in 1:ncol(df.daily)){
  setTxtProgressBar(pb,i)

  tmp.stn = data.frame(DATE=full_dates$DATE, prcp=as.numeric(df.daily[,i]))
  tmp.stn$yr = year(ymd(tmp.stn$DATE))
  tmp.stn$nday = rollsum(tmp.stn$prcp, k=dur, align="right", na.pad=TRUE)
  tmp.out = ddply(tmp.stn, .(yr), summarise, maxp = max(nday,na.rm=TRUE))
  tmp.out$maxp[tmp.out$maxp < 0] = NA
  anmax.out[,i] = as.numeric(tmp.out$maxp)
}
```

You should only save those sites that have at least ~20 years of data, as follows:

```{r eval=FALSE}
# boolean indicating which sites have at least 20 observations
min.id = apply(!is.na(anmax.out), 2, sum) > 19
anmax.out2 = anmax.out[,min.id]
df.meta.out = df.meta[min.id,]

# save it. do NOT save the annual maxima to the data folder
write.csv(anmax.out2, paste0(idir,"atsite-anmax-",dur,"-day.csv"), row.names=FALSE)
write.csv(df.meta.out, paste0(idir,"atsite-anmax-",dur,"-day-metadata.csv"), row.names=FALSE)
```

## PRISM Variables {#prism}

PRISM variables are used to create the trend rasters, which are essentially interpolated GEV parameters. PRISM variables come in 800m and 4km resolution for CONUS, so you want to crop those rasters to the general study area to make them easier to work with. I would recommend using the 4km to avoid blowing up your laptop. 

First, download all the PRISM variables from the website. To do this, navigate to the [30-Year Normals](https://prism.oregonstate.edu/normals/) page. Make sure the `4km` spatial resolution radio button is selected. You want to download every variable for the annual temporal period. Make sure the `annual data` radio button is selected. Then, one by one, select each variable from the drop down menu and download the data in .bil format. Alternatively, you can click `Download All Normals Data (.bil)`. This will save you clicking time but will take up more room on your hard drive. 

```{r}
knitr::include_url("https://prism.oregonstate.edu/normals/")
```

I just went back to the PRISM website while working through this document, and there is no more `Download All Normals Data (.bil)` option. To save yourself the hassle of downloading them all individually, you can download all the PRISM data you need for this analysis here: 

```{r eval=FALSE}
\\CPC-Netapp3.eis.ds.usace.army.mil\RMC\RMC-HECSA\TRANSFER\RMC\Verdin\RMCmsp_v1.0_beta\PRISM_MSP
```

Once all data have been downloaded, you can use the following code (`0d_crop_prism.R`) to crop them to a user-defined bounding box and save them all in an RDS file. 

```{r eval=FALSE}
library(terra)

# set datadir to directory where PRISM was downloaded
datadir = "C:/Path/To/PRISM/Data/"

# load the metadata from the previous step
meta = read.csv(paste0(idir,"/atsite-anmax-",dur,"-day-metadata.csv"))

# these are the raster variables we're going to use - DO NOT MODIFY THESE TWO LINES
raster_vars = c("ppt", "tmean", "tmax", "tmin", "tdmean", "vpdmax", "vpdmin", "solclear", "solslope", "soltotal", "soltrans")
m_vars = c(4, 5, 5, 5, 5, 5, 5, 3, 3, 3, 3)

# read in the first prism file
prism_file = paste0(datadir,"/PRISM_VARIABLE_30yr_normal_800mM5_annual_asc.asc")

# Define the bounding box with extra buffer to avoid NA at stations
lon_min = floor(min(meta$lon))
lon_max = ceiling(max(meta$lon))
lat_min = floor(min(meta$lat))
lat_max = ceiling(max(meta$lat))

# Define the extent of the bounding box
bbox_extent = ext(lon_min, lon_max, lat_min, lat_max)

# loop through all raster variables, modify the string/filename, read and crop
# assign puts the cropped prism data into that variable name for convenience :)
pb = txtProgressBar(min=0, max=length(raster_vars), style=3)
for(i in 1:length(raster_vars)){
  setTxtProgressBar(pb,i)
  
  var = raster_vars[i]
  modified_string <- gsub("VARIABLE", var, prism_file)
  PRISM_data = rast(gsub('5', m_vars[i], modified_string)) 

  PRISM_data <- crop(PRISM_data, bbox_extent)

  assign(var, PRISM_data)
}

# read in the DEM and crop
PRISM_us_dem_800m_r = rast(paste0(datadir,"/PRISM_us_dem_800m_bil.bil"))
PRISM_us_dem_800m_r = crop(PRISM_us_dem_800m_r, bbox_extent)

# put them all in a list and save as RDS file
list.out <- list(Z = PRISM_us_dem_800m_r,
                 ppt = ppt, tmean = tmean, tmax = tmax, tmin = tmin, tdmean = tdmean,
                 vpdmax = vpdmax, vpdmin = vpdmin, solclear = solclear, solslope = solslope,
                 soltotal = soltotal, soltrans = soltrans)
saveRDS(list.out, paste0(idir,"/prism_crop.rds"))

```

## Targets and Target Areas {#targets}

MSP simulations are very computationally expensive. Sensitivities were run by Brian Skahill on the Trinity and Willamette River basins (Skahill et al. 2023) and it was determined that defining some target locations and assigning areas to those locations using Thiessen polygons created basin average precipitation-frequency curves comparable to those when simulating at all pixels. How you define those target locations and areas should be a project specific task, with good representation of orographic influence, topography, and hydroclimatology. The required input for the MSP workflow is a matrix with latitude, longitude, and area for the target locations. To be compatible with these functions, the column names need to be: `lat`, `lon`, and `area`.

The targets are occasionally referred to as subbasin targets because this is one method that could be used. If there is an existing HMS model, the subbasin shapefile can be used to define the target locations as the subbasin centroids, with the area being the total area of each subbasin.

Another option could be equally spaced points across a watershed. The tricky part here is obtaining the corresponding area. For this, a routine such as Thiessen polygons comes in handy. 


## Setting Up Your Workspace

You now have all the required inputs to get started on your MSP analysis! The next chapter will walk you through the four main functions included in the RMCmsp library. See Appendix \@ref(appendixb) for an example workflow.
